---
title: "Peru DHS - BMI trend analysis"
author: "Boris M Fazio"
date: "2017-03-30"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
datafdir  <- "C:/Users/bfazio/Desktop/analisis/modlineal/datafiles/"
#sourcedir <- "C:/Users/bfazio/Desktop/analisis/modlineal/ENDES/"
#scriptdir <- "C:/Users/bfazio/Desktop/analisis/modlineal/rscripts/"
#ISO 8859-1 encoding
```

## Raw data

Load libraries and the aggregated raw data. If you haven't generated the latter, first do so using [this function](https://github.com/bmfazio/dhs-analysis/blob/master/peru/genfile-BMI.R).

```{r warning=FALSE}
library(magrittr)
library(data.table)

dhs <- fread(datafdir %>% paste0("rawdata.csv"))
str(dhs) #Always good to take a look.
```

## Data cleaning and preparation

For this analysis, the relationship of ethnicity and of residence history with Body Mass Index (BMI) will be explored over the years 2005 to 2015. We first need to construct those variables from the available data.

### Variable construction

#### Ethnicity

Following the operationalization proposed in (Valdivia 2007), we create three categories (old spanish speakers, recent spanish speakers and indigenous language speakers) based on the following questions asked in the Peru DHS:

* `ethnic_self`: ¿Qué idioma o dialecto hablan habitualmente en su hogar? (*What language or dialect do you normally speak at home?*)
* `ethnic_mom`: ¿Cuál es el idioma o lengua materna de su madre? (*What is your mother's mother tongue?*)
* `ethnic_dad`: ¿Cuál es el idioma o lengua materna de su padre? (*What is your father's mother tongue?*)

Valid answers are *spanish*, *quechua*, *aymara*, *other indigenous language* and *foreign language*.

This set of questions was not asked during before 2005 or in the 2009-2012 surveys, so we exclude those years from this analysis.

```{r}
#ethnic_self, ethnic_dad, ethnic_mom
#1:spanish, 2:quechua, 3:aymara, 4:other indigenous, 5:foreign

dhs[ethnic_self==1&ethnic_dad==1&ethnic_mom==1,ethnic_cat:="old spanish"]
dhs[ethnic_self==1&(is.element(ethnic_dad,2:4)|is.element(ethnic_mom,2:4)),ethnic_cat:="recent spanish"]
dhs[is.element(ethnic_self,2:4),ethnic_cat:="indigenous"]

dhs[!(year%>%is.element(c(2003,2004,2009:2012)))] %>% with(ethnic_cat %>% table(year))
```

Notice we categorize respondents as `"indigenous"` without using their parent's data. From 2013 onwards, a question skip was introduced whenever the respondent says they use an indigenous language at home. This is not an issue because almost all indigenous language users had indigenous language-speaking parents. We can examine the 2005-2008 data to see this:

```{r}
dhs[is.element(year,2005:2008)] %T>% with(table(ethnic_self,ethnic_mom) %>% print) %>% with(table(ethnic_self,ethnic_dad))
```

#### Residence history

This variable will have three categories based on the respondent's current place of residence, their previous place of residence (if any) and their childhood place of residence: lifetime urban dweller, lifetime rural dweller and rural-to-urban migrant.

```{r}
#res_ur
#1:urban, 2:rural
#res_child, res_prev
#0:capital/large city, 1:city, 2:town, 3:countryside, 4:abroad

dhs[res_ur==1&is.element(res_child,0:2)&(is.element(res_prev,0:2)|is.na(res_prev)),residence_cat:="urban"]
dhs[res_ur==2&res_child==3&(res_prev==3|is.na(res_prev)),residence_cat:="rural"]
dhs[res_ur==1&res_child==3&(res_prev==3|!is.na(res_prev)),residence_cat:="migrant"]

dhs %>% with(residence_cat %>% table(year))
```

### Case exclusion

We perform the following exclusions from our dataset:

* Respondents that don't identify as residents of the household
* Respondents that report being pregnant at the time of the survey
* Respondents that fail to be described by any of the variables constructed above

```{r}
dhs <- dhs[qc_pregnant==0&qc_visitor==1&
             !(is.na(ethnic_cat)&is.na(residence_cat))]
```

Because `bmi` is the dependent variable, we must exclude any cases without the measurement but we also need to check that doing so may inadvertently bias our data. **This check is probably too simplistic but I'm not sure what else to do right now. Probably need to use sample weights at some point here.**

```{r}
dhs %T>%
  with(table(year,isNA=(is.na(bmi)|bmi==9999)) %>% prop.table(margin=1) %>% '*'(100) %>% round(2) %>% print) %T>%
  with(table(ethnic_cat,isNA=(is.na(bmi)|bmi==9999)) %>% prop.table(margin=1) %>% '*'(100) %>% round(2) %>% print) %>%
  with(table(residence_cat,isNA=(is.na(bmi)|bmi==9999)) %>% prop.table(margin=1) %>% '*'(100) %>% round(2) %>% print)

dhs <- dhs[!(is.na(bmi)|bmi==9999)] #Drop it!
```

As it turns out, no anthropometric measurements were taken during the 2006 survey. Bummer, we'll just have to have a gap in the trend.

### Finishing touches

Finally, let's make sure all categorical values are labeled and drop variables that we won't be using anymore. We're also taking the `education` variable into a more commonly used form by collapsing some of its levels **(technically you never should do this without good reason... well... just move along k?)**.

```{r}
dhs[, c("cluster","hhnumber","hhline","qc_pregnant","qc_visitor","res_ur","res_type","res_child","res_prev","ethnic_mom","ethnic_dad","ethnic_self") := NULL]

dhs[,age5cat:=factor(x=age5cat,labels=c("15-19","20-24","25-29","30-34","35-39","40-44","45-49"),ordered=TRUE)]
dhs[is.element(education,c(1,3)),education:=(education-1L)]
dhs[,education:=factor(x=education,labels=c("Did not complete primary","Only completed primary","Complete secondary","Higher education"),ordered=TRUE)]
dhs[,region:=factor(x=region,labels=c("Amazonas","Ancash","Apurimac","Arequipa","Ayacucho","Cajamarca","Callao","Cusco","Huancavelica","Huanuco","Ica","Junin","La Libertad","Lambayeque","Lima","Loreto","Madre de Dios","Moquegua","Pasco","Piura","Puno","San Martin","Tacna","Tumbes","Ucayali"),ordered=TRUE)]
dhs[is.element(res_time,96:99),res_time:=NA]
dhs[,wealth:=factor(x=wealth,labels=c("poorest","poorer","middle","richer","richest"),ordered=TRUE)]
dhs[,natregion:=factor(x=natregion,labels=c("Lima metropolitana","Resto costa","Sierra","Selva","Selva baja"),ordered=TRUE)]
```

## Dataset ready for analysis!

Let's take a final look at our reproducibly-prepared data.

```{r pressure, echo=FALSE}
print(str(dhs))
print(dhs)
```
